---
title: "SARS-CoV-2 vs Influenza: XGBoost Classifier Model"
author: "Patrick G. Lyons, MD, MSc"
output: html_document
date: "2022-08-26"
---

This script builds an XGBoost classifer model to predict death/hospice discharge in SARS-CoV-2 and influenza patients.

# Load libraries
```{r libraries, include = F}
library(doParallel)
library(data.table)
library(tidymodels)
library(tidytable)
library(tidyverse)
library(janitor)
library(here)
```

# Multicore processing
``` {r multicore, include = F}
all_cores <- parallel::detectCores(logical = FALSE)
ps_clustr <- parallel::makePSOCKcluster(all_cores-1)
doParallel::registerDoParallel(ps_clustr)
```

# Load data
```{r load data}
df <- fread(here("covflu_combined_deidentified.csv"))
```

# Stratify data into two groups based on virus
```{r stratify data}
cov <- df %>%
  filter.(virus == "SARS-CoV-2") %>%
  select.(-virus)

flu <- df %>%
  filter.(virus == "Influenza") %>%
  select.(-virus)
```

# Build logistic regression model for each virus and calculate AUROC
```{r logsitic regression}
# COVID model
model_c <-
  glm(
    dead_or_hospice ~ 
      age +
      bmi + 
      female + 
      min_albumin +
      max_ast + 
      max_o2_flow + 
      max_anc +
      max_abs_lymph +
      min_hcrt +
      min_sodium +
      max_resp +
      min_spo2 +
      min_platelet,
    family = "binomial",
    data   = cov
  )

# influenza model
model_f <-
  glm(
    dead_or_hospice ~ 
      age +
      bmi + 
      min_albumin +
      min_hcrt +
      max_pulse +
      max_temp +
      max_rdw +
      min_platelet,
    family = "binomial",
    data   = cov
  )

# predict death/hospice in COVID patients using COVID model
prob_cc <- predict(model_c, newdata = cov, type = "response")
# predict death/hospice in influenza patients using COVID model
prob_cf <- predict(model_c, newdata = flu, type = "response")
# predict death/hospice in influenza patients using COVID model
prob_ff <- predict(model_f, newdata = flu, type = "response")
# predict death/hospice in COVID patients using influenza model
prob_fc <- predict(model_f, newdata = cov, type = "response")

# calculate AUROC values
pROC::auc(pROC::roc(cov$dead_or_hospice, prob_cc))
pROC::auc(pROC::roc(flu$dead_or_hospice, prob_cf))
pROC::auc(pROC::roc(flu$dead_or_hospice, prob_ff))
pROC::auc(pROC::roc(cov$dead_or_hospice, prob_fc))
```

# XGBoost in split train/test datasets
```{r recipes}
# set model specifications
xgb_spec <- 
  boost_tree(
    trees          = 1000, 
    tree_depth     = tune(), 
    min_n          = tune(), 
    loss_reduction = tune(),                     
    sample_size    = tune(), 
    mtry           = tune(),         
    learn_rate     = tune(),                         
  ) %>% 
  set_engine("xgboost") %>% 
  set_mode("classification")

# set COVID tuning grid
xgb_grid_cov <- 
  grid_max_entropy(
    tree_depth(),
    min_n(),
    loss_reduction(),
    sample_size = sample_prop(),
    finalize(mtry(), cov),
    learn_rate(),
    size = 60
  )

# set influenza tuning grid
xgb_grid_flu <- 
  grid_max_entropy(
    tree_depth(),
    min_n(),
    loss_reduction(),
    sample_size = sample_prop(),
    finalize(mtry(), flu),
    learn_rate(),
    size = 60
  )

xgb_wf <- 
  workflow() %>%
  add_formula(dead_or_hospice ~ .) %>%
  add_model(xgb_spec)

# tuning bootstraps
cov <- cov %>% mutate.(dead_or_hospice = factor(dead_or_hospice))
flu <- flu %>% mutate.(dead_or_hospice = factor(dead_or_hospice))

set.seed(5432)

# COVID training/testing split
cov_split <- initial_split(cov, strata = dead_or_hospice)
cov_train <- training(cov_split)
cov_tests <- testing(cov_split)

# influenza training/testing split
flu_split <- initial_split(flu, strata = dead_or_hospice)
flu_train <- training(flu_split)
flu_tests <- testing(flu_split)

set.seed(2345)
cov_boots <- cov_train %>% bootstraps(50, strata = dead_or_hospice, apparent = TRUE)
flu_boots <- flu_train %>% bootstraps(50, strata = dead_or_hospice, apparent = TRUE)
```

# Model tuning
```{r tuning}
grid <- 
  control_grid(
    allow_par     = TRUE, 
    parallel_over = "resamples",
    save_pred     = TRUE
  )

# save tuned grid for COVID
tuned_cov <- 
  tune_grid(
    xgb_wf,
    resamples = cov_boots,
    grid      = xgb_grid_cov,
    control   = grid
  )

# save tuned grid for influenza
tuned_flu <- 
  tune_grid(
    xgb_wf,
    resamples = flu_boots,
    grid      = xgb_grid_flu,
    control   = grid
  )

# save highest performing workflows for each virus
final_wf_cov <- finalize_workflow(xgb_wf, select_best(tuned_cov, "roc_auc"))
final_wf_flu <- finalize_workflow(xgb_wf, select_best(tuned_flu, "roc_auc"))
```

Internal testing
```{r xgboost interval validation}
# create new bootstraps for internal validation
set.seed(123)
cboots <- cov_train %>% bootstraps(1000, strata = dead_or_hospice, apparent = TRUE)
fboots <- cov_train %>% bootstraps(1000, strata = dead_or_hospice, apparent = TRUE)

# fit new COVID bootstraps to COVID model
final_cov <- 
  fit_resamples(
    final_wf_cov, 
    resamples = cboots, 
    control   = control_resamples(save_pred = T)
  )

# fit new influenza bootstraps to influenza model
final_flu <- 
  fit_resamples(
    final_wf_flu, 
    resamples = fboots, 
    control   = control_resamples(save_pred = T)
  )

# combine final COVID model with final influenza model
x <-
  final_cov %>% mutate.(virus = "SARS-CoV-2") %>%
  bind_rows.(final_flu %>% mutate.(virus = "Influenza")) %>%
  nest_by.(virus)

# AUROC, 95% confidence intervals for each virus
x %>%
  unnest.(data) %>%
  unnest.(.metrics) %>%
  filter.(.metric == "roc_auc") %>%
  summarize.(
    auroc_mean = mean(.estimate, na.rm = TRUE),
    ci_low     = quantile(.estimate, probs = 0.025, na.rm = TRUE),
    ci_high    = quantile(.estimate, probs = 0.975, na.rm = TRUE),
    .by        = c(virus)
  ) 
```

# Use testing set data on models
```{r xgboost external validation}
# create function to fit model and save AUROC
preds_fn <- function(split, fit) {
  pred <- predict(fit, new_data = assessment(split), type = "prob")
  pred %>%
    bind_cols.(assessment(split)) %>%
    roc_auc(dead_or_hospice, .pred_0) %>%
    pull(.estimate)
}

# fit training data on model
fit_cov <- fit(final_wf_cov, data = cov_train)
fit_flu <- fit(final_wf_flu, data = flu_train)

# create 1000 bootstrap samples from testing sets
cboots2 <- cov_tests %>% bootstraps(1000, strata = dead_or_hospice, apparent = T)
fboots2 <- flu_tests %>% bootstraps(1000, strata = dead_or_hospice, apparent = T)

# find AUROC and 95% confidence intervals for both models
cboots2 %>%
  mutate.(auroc = map_dbl.(splits, preds_fn, fit_cov),) %>%
  summarize.(
    auroc_mean  = mean(auroc, na.rm = TRUE),
    ci_low      = quantile(auroc, probs = 0.025, na.rm = TRUE),
    ci_high     = quantile(auroc, probs = 0.975, na.rm = TRUE)
  )

fboots2 %>%
  mutate.(auroc = map_dbl.(splits, preds_fn, fit_flu),) %>%
  summarize.(
    auroc_mean  = mean(auroc, na.rm = TRUE),
    ci_low      = quantile(auroc, probs = 0.025, na.rm = TRUE),
    ci_high     = quantile(auroc, probs = 0.975, na.rm = TRUE)
  )
```